---
title: "Figure 1: hypothesis tests for seven LLM models"
format: html
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)
```

# Reports (combined automated evaluation strategy)

This document repeats the five automated-evaluation one-way ANOVA tests from
`figure_1.qmd` after combining its three-model data with the four-model data
used by `figure_1_Aug23.qmd`. The human-evaluation data are not included because
they contain observations for only four of the seven models.

```{r}
#| label: combine-data

old_case_df <- read.csv(
  "data/token_analysis/automatic_case_level_metrics.csv",
  row.names = 1
)

new_case_df <- read.csv(
  "data/token_analysis/automatic_newmodel_case_level_metrics.csv",
  row.names = 1
)

metric_names <- c(
  "avg_word_length",
  "total_words",
  "unique_tokens",
  "avg_sentence_length",
  "total_sentences"
)

required_columns <- c("model", "case_id", metric_names)
stopifnot(
  all(required_columns %in% names(old_case_df)),
  all(required_columns %in% names(new_case_df))
)

case_df <- bind_rows(
  old_case_df %>% select(all_of(required_columns)),
  new_case_df %>% select(all_of(required_columns))
)

model_order <- c(
  "claude-3-7-sonnet-20250219",
  "gemini-2.0-flash",
  "gpt-4o",
  "claude-opus-5",
  "gemini-3.6-flash",
  "gpt-5.6-sol",
  "o1-2024-12-17"
)

stopifnot(
  setequal(unique(case_df$model), model_order),
  !anyDuplicated(case_df[c("model", "case_id")]),
  !anyNA(case_df[metric_names])
)

case_df$model <- factor(case_df$model, levels = model_order)

model_counts <- case_df %>% count(model, name = "n_cases")
print(model_counts)
stopifnot(nrow(model_counts) == 7L, length(unique(model_counts$n_cases)) == 1L)

df_long <- case_df %>%
  pivot_longer(
    cols = all_of(metric_names),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = factor(metric, levels = metric_names))

model_colors <- c(
  "#EF6F6A", "#6388B4", "#55AD89", "#F4A09C",
  "#93ACCB", "#E6B655", "#64CDCC"
)
```

The null hypothesis for each test is that the mean metric is equal across all
seven models. The alternative is that at least one model mean differs.

```{r}
#| label: plot-function

plot_metric <- function(data, y_label) {
  ggplot(data, aes(x = model, y = value, fill = model)) +
    geom_boxplot() +
    scale_fill_manual(values = model_colors, drop = FALSE) +
    labs(x = NULL, y = y_label) +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
}
```

## Average word length

```{r}
tmp <- filter(df_long, metric == "avg_word_length")
summary(aov(value ~ model, data = tmp))
plot_metric(tmp, "Average word length")
```

## Total words

```{r}
tmp <- filter(df_long, metric == "total_words")
summary(aov(value ~ model, data = tmp))
plot_metric(tmp, "Total words")
```

## Unique tokens

```{r}
tmp <- filter(df_long, metric == "unique_tokens")
summary(aov(value ~ model, data = tmp))
plot_metric(tmp, "Unique tokens")
```

## Average sentence length

```{r}
tmp <- filter(df_long, metric == "avg_sentence_length")
summary(aov(value ~ model, data = tmp))
plot_metric(tmp, "Average sentence length")
```

## Total sentences

```{r}
tmp <- filter(df_long, metric == "total_sentences")
summary(aov(value ~ model, data = tmp))
plot_metric(tmp, "Total sentences")
```
